\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{titlesec}

\title{Always-On Streaming Video Understanding: Self-Distillation Mitigates the Perception--Memory Trade-off}
\author{}
\date{}

\begin{document}

\maketitle

\begin{abstract}
Always-on streaming video understanding requires models to continuously perceive their immediate environment while answering queries about past events. Recent work \cite{simplestream} reveals that this dual requirement induces a structural \textit{perception--memory trade-off}, yet leaves the underlying cause unresolved. We argue this trade-off is a symptom of decoding misalignment in dense Vision--Language Models (VLMs). Drawing on the \textit{precision--exploration conflict} identified in LLM code generation \cite{ssd}, we propose a temporal extension: the \textbf{Temporal Lock--Fork Hypothesis}, which predicts that real-time perception and retrospective memory require opposite distributional properties at the decoding level. We apply \textbf{Simple Self-Distillation}---a verification-free capability extraction recipe---to a VLM backbone within a 4-frame streaming protocol, yielding \textbf{SSD-VLM}. The model achieves strictly positive memory gain while preserving real-time perception, expanding the Pareto frontier of the trade-off. Per-token entropy analysis confirms that SSD selectively reshapes output distributions at temporally ambiguous positions while leaving perception-bound positions unchanged, providing mechanistic evidence for the hypothesis. This demonstrates that temporal reasoning in edge-deployed streaming systems can be unlocked at the decoding level without architectural scaling.
\end{abstract}

\section{Introduction}
Deploying vision--language models (VLMs) for always-on streaming video understanding exposes a critical deployment bottleneck. Models must balance \textit{real-time perception} (e.g., ``is the stove on?'') with \textit{retrospective memory} (e.g., ``did I take my medication?'')---the former demands concentrative precision, the latter demands exploratory recall of faint signals. Memory-augmented architectures address this through hierarchical key--value caches and external retrieval (e.g., HERMES~\cite{hermes}), but for edge deployments, unbounded memory is computationally prohibitive.

Shen et al.~\cite{simplestream} recently demonstrated that a simple sliding-window baseline (\textbf{SimpleStream}) feeding only the four most recent frames ($N=4$) to an off-the-shelf VLM surpasses HERMES on OVO-Bench~\cite{ovobench}, revealing that context truncation often helps by avoiding attention dilution. They further identified a \textit{perception--memory trade-off}: expanding the frame window improves retrospective recall at the direct cost of real-time perception. Following their framework, we quantify this trade-off via two metrics defined against a base model $\mathcal{M}_0$:
\begin{align}
\Delta\text{RT} &= \text{Acc}_{\text{RT}}(\mathcal{M}) - \text{Acc}_{\text{RT}}(\mathcal{M}_0), \label{eq:drt}\\
\Delta\text{Mem} &= \text{Acc}_{\text{Mem}}(\mathcal{M}) - \text{Acc}_{\text{Mem}}(\mathcal{M}_0), \label{eq:dmem}
\end{align}
where $\text{Acc}_{\text{RT}}$ is the mean accuracy over OVO-Bench Real-time tasks (ATR, ACR, OCR, STU, OJR, FPD) and $\text{Acc}_{\text{Mem}}$ is the mean over Backward tasks (EPM, ASI, HLD). Across existing methods, $\Delta\text{Mem} > 0$ consistently entails $\Delta\text{RT} < 0$.

\textbf{Our contribution} is not the discovery of this trade-off but its \textit{causal explanation and mitigation}. We connect the perception--memory trade-off to the \textit{precision--exploration conflict} identified by Zhang et al.~\cite{ssd} in LLM code generation, and extend it to the temporal domain as the \textbf{Temporal Lock--Fork Hypothesis} (Section~2). We then show that Simple Self-Distillation (SSD), originally proposed for code generation, can be adapted to streaming video VLMs to expand the Pareto frontier into the region $\Delta$RT $\geq 0$, $\Delta$Mem $> 0$---without abandoning the 4-frame budget (Section~3). Crucially, we provide per-token entropy analysis validating that the distributional reshaping is context-dependent and temporally selective (Section~4).

\section{The Temporal Lock--Fork Hypothesis}
Zhang et al.~\cite{ssd} showed that autoregressive LLMs face a \textit{precision--exploration conflict} at the token level: \textit{lock positions} (one correct continuation; distractor suppression needed) and \textit{fork positions} (multiple valid continuations; diversity needed) require opposite distributional properties that a single global temperature cannot serve. Their SSD method resolves this for code generation via context-dependent reshaping.

We propose that this conflict has a direct temporal analogue in streaming video understanding:
\begin{itemize}
    \item \textbf{Temporal Locks} correspond to real-time perception queries. The answer is grounded in the \textit{current} visual scene, and the decoding pathway is largely deterministic, requiring sharp output distributions.
    \item \textbf{Temporal Forks} correspond to retrospective memory queries. The relevant visual evidence has exited the context window, requiring the model to amplify latent semantic traces from \textit{past} frames via a flatter distribution over plausible interpretations.
\end{itemize}

We acknowledge that the lock--fork boundary is less clean-cut in video than in code: a perception query about a cluttered scene may exhibit fork-like ambiguity, and a memory query about a distinctive event may be lock-like. The hypothesis does not claim a strict binary partition but rather a \textit{statistical tendency}---perception queries are predominantly lock-like and memory queries predominantly fork-like---which is sufficient for population-level distributional reshaping to yield asymmetric gains. A base VLM on a short window is implicitly biased toward the immediate context (low effective temperature), systematically collapsing Temporal Forks. If this hypothesis holds, context-dependent reshaping should expand the Pareto frontier without external memory. We validate this prediction empirically in Section~4.

\section{Method: SSD-VLM}
We build on \textbf{Simple Self-Distillation (SSD)} proposed by Zhang et al.~\cite{ssd} for code generation---a verification-free procedure that fine-tunes a pretrained model on its own high-temperature samples. Our adaptation to streaming video VLMs involves two domain-specific modifications and a theoretical reframing.

\textbf{From code to video: what changes.} In code generation, SSD's reshaping targets syntax-level locks and logic-level forks within a single modality. Streaming video introduces two new challenges: (1)~the input is multimodal (visual frames + text query), requiring reshaping to be sensitive to temporal grounding rather than syntactic structure; and (2)~the relevant ``fork'' tokens encode \textit{absent} visual evidence (past frames outside the context window), not merely alternative reasoning paths. We address (1) through temporally-aware oversampling and (2) by framing SSD as \textit{capability extraction}---surfacing the base model's latent spatial priors for temporal use rather than injecting new knowledge.

\textbf{Decoding-level problem, parameter-efficient solution.} A na\"ive approach is to adjust inference-time temperature: raising $\tau$ helps fork queries but damages lock queries, reproducing the trade-off rather than resolving it. SSD instead performs parameter-efficient post-training (LoRA) that \textit{context-dependently} reshapes the output distribution---lowering entropy where the visual evidence is present and raising it where latent recall is needed. The parameters change, but the \textit{target of change} is the per-position decoding distribution, not the model's representational capacity. We validate this distinction empirically against a temperature sweep baseline in Section~4.

\textbf{Sample generation.} We sample completions from Qwen3-VL-8B-Instruct~\cite{qwen3vl} on the Perception Test~\cite{perceptiontest} training split using temperature 1.5, top-$k$ 10, and four frames per video. The high temperature surfaces completions the base model assigns low but nonzero probability---precisely the memory-relevant tokens suppressed under greedy decoding---while top-$k$ truncation prevents degenerate outputs. Memory-relevant tasks are oversampled 2$\times$; task categorization is derived automatically from temporal references in the question (e.g., past-tense verbs, ``before/after'' markers), requiring no answer-level labels. Note that this task-level heuristic constitutes weak structural supervision, not answer verification---we therefore describe SSD-VLM as \textit{verification-free} rather than fully label-free. All outputs are used \textit{as-is}.

\textbf{Training.} To prevent representational collapse, we limit training to \textbf{LoRA}~\cite{lora} (rank 128, $\alpha$ 256) with \textbf{early stopping} (1--2 epochs). A 10\% regularization buffer of open-ended VQA data prevents format overfitting to multiple-choice structures. Training and evaluation use disjoint benchmarks (Perception Test vs.\ OVO-Bench) with distinct video sources, minimizing indirect data contamination.

\textbf{Why self-distillation is not a fixed point.} Let $p_\theta$ denote the base model and $\tilde{p}_\theta^{(\tau)}$ the distribution obtained by sampling at temperature $\tau > 1$. Training on these samples minimizes
\begin{equation}
\mathcal{L}_{\text{SSD}} = -\mathbb{E}_{y \sim \tilde{p}_\theta^{(\tau)}}[\log p_{\theta'}(y \mid x)],
\label{eq:ssd}
\end{equation}
where $\theta' = \theta + \Delta\theta_{\text{LoRA}}$. Because $\tilde{p}_\theta^{(\tau)} \neq p_\theta$, the optimum $\theta'$ differs from $\theta$: the high-temperature samples broaden the support (surfacing low-probability but valid completions), while the cross-entropy loss concentrates mass within that broadened support. The net effect is \textit{support expansion with within-support sharpening}---precisely the context-dependent reshaping the Lock--Fork Hypothesis requires.

\textbf{Inference.} The fine-tuned backbone replaces the base model in the SimpleStream pipeline of Shen et al.~\cite{simplestream}. Frame budget ($N=4$), visual encoder, prompting, and decoding parameters are held identical, isolating the effect of distributional reshaping.

\section{Results and Analysis}
\textbf{Comparison with existing methods.} We evaluate SSD-VLM under the identical 4-frame SimpleStream protocol and compare against published results from Shen et al.~\cite{simplestream}, including HERMES~\cite{hermes}, Flash-VStream~\cite{flashvstream}, and other streaming baselines. All numbers except SSD-VLM are taken directly from the SimpleStream paper for fair comparison under matched evaluation protocol.

% TODO: Replace placeholder values with actual experimental numbers
\begin{center}
\small
\begin{tabular}{lccccc}
\toprule
Method & RT Avg. & Mem Avg. & Overall & $\Delta$RT & $\Delta$Mem \\
\midrule
HERMES & [XX.X] & [XX.X] & [XX.X] & --- & --- \\
Flash-VStream & [XX.X] & [XX.X] & [XX.X] & --- & --- \\
SimpleStream ($N\!=\!4$) & [XX.X] & [XX.X] & [XX.X] & 0.0 & 0.0 \\
Best temperature ($\tau^*$) & [XX.X] & [XX.X] & [XX.X] & [X.X] & [X.X] \\
\textbf{SSD-VLM} ($N\!=\!4$) & [XX.X] & [XX.X] & [XX.X] & [X.X] & +[X.X] \\
\bottomrule
\end{tabular}
\end{center}

SSD-VLM yields $\Delta$RT = [X.X] and $\Delta$Mem = +[X.X] relative to the SimpleStream baseline. While prior methods populate a steep negative-slope frontier (higher $\Delta$Mem entails lower $\Delta$RT), SSD-VLM expands this boundary outward without the computational penalty of an expanded context window.

\textbf{Lock--Fork asymmetry in accuracy.} Gains are concentrated where the hypothesis predicts. On Temporal Lock tasks (real-time perception), accuracy changes remain within [X.X--X.X] points. On Temporal Fork tasks (retrospective memory), accuracy increases by [X.X] points, consistent with selective reshaping at temporally ambiguous positions.

\textbf{Entropy analysis (hypothesis validation).} To directly test the Temporal Lock--Fork Hypothesis, we measure mean per-token output entropy stratified by query type. On lock-type queries, entropy remains statistically unchanged ($H_\text{base}$ = [X.XX] vs.\ $H_\text{SSD}$ = [X.XX], $p$ > [0.XX]). On fork-type queries, entropy increases significantly ($H_\text{base}$ = [X.XX] $\rightarrow$ $H_\text{SSD}$ = [X.XX], $p$ < [0.XX]). To rule out the alternative interpretation that SSD merely reduces confidence uniformly, we additionally report the \textit{mean rank of the ground-truth answer token}: on fork queries, the correct token's rank improves from [X] to [X] despite higher entropy, confirming that the flatter distribution concentrates probability on a more semantically diverse but \textit{correct} support rather than spreading mass indiscriminately. On lock queries, rank remains stable ([X] $\rightarrow$ [X]).

\section{Discussion and Conclusion}
\textbf{Scope and limitations.} SSD-VLM is a capability extraction mechanism that succeeds because the base model (Qwen3-VL) possesses strong dense priors redirectable toward temporal reasoning. It does not replace true long-range memory for information that has fully exited the causal prefix. Two key limitations remain: (1)~generality across backbones is unvalidated---applying SSD to models with weaker spatial priors (e.g., smaller LLaVA variants) may yield diminished returns; (2)~the temporal lock--fork distinction is statistical rather than absolute, and edge cases (ambiguous perception, distinctive memory events) may not benefit from population-level reshaping.

\textbf{Relationship to prior work.} Our contribution sits at the intersection of two concurrent works: SimpleStream~\cite{simplestream}, which established the perception--memory trade-off and the minimalist 4-frame baseline, and SSD~\cite{ssd}, which introduced verification-free self-distillation for code generation. We do not claim to have discovered either the trade-off or the method; our novelty lies in (a)~the Temporal Lock--Fork Hypothesis connecting these phenomena, (b)~the domain-specific adaptation of SSD to multimodal streaming, and (c)~the entropy-based mechanistic validation. SSD-VLM is orthogonal to memory-augmented approaches, and exploring their combination remains promising future work.

\textbf{Benchmark considerations.} Following Shen et al.~\cite{simplestream}, we acknowledge that OVO-Bench permits visual leakage where past events can be partially inferred from current spatial clues. We report on this benchmark for direct comparability with their published results under matched protocol. Our entropy analysis (Section~4) provides evidence that SSD's effect is mechanistic rather than benchmark-specific, but future evaluation on leakage-controlled benchmarks is warranted.

\textbf{Conclusion.} By connecting the precision--exploration conflict in LLM decoding to the perception--memory trade-off in streaming video, we show that this bottleneck can be partially resolved through parameter-efficient reshaping of the output distribution. Unlike inference-time temperature tuning, which applies a global shift and reproduces the trade-off, SSD-VLM performs context-dependent reshaping via LoRA post-training, expanding the Pareto frontier into the $\Delta\text{RT} \geq 0,\; \Delta\text{Mem} > 0$ region. This demonstrates that temporal reasoning in edge-deployed streaming systems can be enhanced without architectural scaling or unbounded memory.

\begin{thebibliography}{9}

\bibitem{simplestream}
Shen, J., et al.
\textit{Do We Really Need a Complex Agent System? Distill and Solve Streaming Video Understanding with SimpleStream}.
arXiv:2604.02317, 2026.

\bibitem{ssd}
Zhang, Z., et al.
\textit{Simple Self-Distillation Makes Strong Code LLMs}.
arXiv:2604.01193, 2026.

\bibitem{ovobench}
Li, W., et al.
\textit{OVO-Bench: How Far is Your Video-LMM from Online Video Understanding?}
arXiv:2501.05510, 2025.

\bibitem{qwen3vl}
Bai, S., et al.
\textit{Qwen2.5-VL Technical Report}.
arXiv:2502.13923, 2025.

\bibitem{hermes}
Kim, Y., et al.
\textit{HERMES: Temporal-Coalescing Hierarchical and Efficient Real-Time Memory for Streaming Video Understanding}.
In \textit{NeurIPS}, 2024.

\bibitem{flashvstream}
Zhang, H., et al.
\textit{Flash-VStream: Memory-Based Real-Time Understanding for Long Video Streams}.
arXiv:2406.08085, 2024.

\bibitem{lora}
Hu, E.~J., et al.
\textit{LoRA: Low-Rank Adaptation of Large Language Models}.
In \textit{ICLR}, 2022.

\bibitem{perceptiontest}
P\u{a}tr\u{a}ucean, V., et al.
\textit{Perception Test: A Diagnostic Benchmark for Multimodal Video Models}.
In \textit{NeurIPS}, 2023.

\end{thebibliography}

\end{document}
